---
title: "Laith Barakat 7042 Project Part 1"
author: "Laith Barakat"
date: "January 21, 2020"
output: word_document
---


##Initial exploration of the data

###Step 1

Read in the files and load relevant libraries

```{r warning=FALSE, message=FALSE}
library(readxl)
library(tidyverse)
setwd('C:/Users/Laith/Documents/SpringSem20/Stat Modeling')
FAA1 <- read_xls('FAA1-1.xls')
FAA2 <- read_xls('FAA2-1.xls')
```


###Step 2

Check structure

```{r}
str(FAA1)
str(FAA2)
```

The data set FAA1 has 800 observations and 8 columns. The dataset FAA2 has 150 variables and 7 columns. FAA1 has all of the same fields as FAA2 with the addition of duration. 

###Step 3

Merging datasets

```{r}
FAAm <- full_join(FAA1, FAA2)
distinct(FAAm)

```

There were duplicates; this is covered by the `distinct()` function in R and will be eliminated.

###Step 4

Structure of combined dataset and summary statistics

```{r}
str(FAAm)
glimpse(FAAm)
summary(FAAm)
```

###Step 5

Insights:

* After I remove duplicates in the dataset and merge together, I have a sample size of 850 flights that will provide a statistically significant sample for our analysis.

* The 642 NA values in the column for air speed imply that the column may be difficult to find significant.

* The minimum value for duration, 14.76, is an outlier that we will have to eliminate.

* There are upper and lower ground speed outliers within that variable. 

* There is a negative value in the height variable. This is an outlier as well. 

##Data Cleaning and Further exploration

###Step 6

Abnormal values

```{r}
sum(FAAm$duration < 40,na.rm = TRUE)
sum(FAAm$speed_ground < 30, na.rm = TRUE)
sum(FAAm$speed_ground > 140, na.rm = TRUE)
sum(FAAm$speed_air < 30, na.rm = TRUE)
sum(FAAm$speed_air > 140, na.rm = TRUE)
sum(FAAm$height < 6, na.rm = TRUE)

FAAmc <- FAAm %>% 
  filter(
    duration > 40 | is.na(duration),
    speed_ground > 30 & speed_ground < 140 | is.na(speed_ground),
    speed_air > 30 & speed_air < 140| is.na(speed_air),
    height > 6 | is.na(height)
  )

nrow(FAAm) - nrow(FAAmc)
```

I ended up removing about 20 rows. 

###Step 7

Repeating step 4

```{r}
str(FAAmc)
glimpse(FAAmc)
summary(FAAmc)
```

###Step 8 

Histograms of each variable

```{r}
par(mfrow = c(2,4))
hist(FAAmc$duration)
hist(FAAmc$no_pasg)
hist(FAAmc$speed_ground)
hist(FAAmc$speed_air)
hist(FAAmc$height)
hist(FAAmc$pitch)
hist(FAAmc$distance)

```

Many of these look to be quite normally distributed.

###Step 9

* Each of the numeric variables, when cleaned and visualized, look normally distributed or relatively normally distributed.

* All values are within the data dictionary's parameters and include 832 observations of data.

* I kept the observations where distance > 6000 because that variable is the response variable.

##Initial analysis for identifying important factors that impact the response variable "landing distance"

###Step 10

Pairwise correlation and creation of Table 1

```{r}
table1 <- matrix(data = c(names(FAAmc[,2:8]), cor(FAAmc[,2:8], use = 'pairwise')[,7], cor(FAAmc[,2:8], use = 'pairwise')[,7] >= 0), nrow = 7, ncol = 3)
table1
```

###Step 11

Scatterplot pairs 

```{r}
pairs(FAAmc[,2:8])
```

Correlation strength in these plots is consistent with the values computed in Step 10.

###Step 12

Including plane as a factor.

```{r}
FAAmc$aircraft <- if_else(FAAmc$aircraft == 'boeing', 1, 0)
table1_1 <- matrix(data = c(names(FAAmc[,1:8]), cor(FAAmc[,1:8], use = 'pairwise')[,8], cor(FAAmc[,1:8], use = 'pairwise')[,8] >= 0), nrow = 8, ncol = 3)
table1_1
cor(FAAmc[,1:8])[,7]
pairs(FAAmc[,1:8])
```


##Regression using a single factor each time

###Step 13

Single-variable regressions and table 1 building

```{r}
lm_aircraft <- lm(distance ~ aircraft, data = FAAmc)
lm_duration <- lm(distance ~ duration, data = FAAmc)
lm_psg <- lm(distance ~ no_pasg, data = FAAmc)
lm_gs <- lm(distance ~ speed_ground, data = FAAmc)
lm_as <- lm(distance ~ speed_air, data = FAAmc)
lm_ht <- lm(distance ~ height, data = FAAmc)
lm_pitch <- lm(distance ~ pitch, data = FAAmc)

lmp <- function (modelobject) {
    if (class(modelobject) != "lm") stop("Not an object of class 'lm' ")
    f <- summary(modelobject)$fstatistic
    p <- pf(f[1],f[2],f[3],lower.tail=F)
    attributes(p) <- NULL
    return(p)
}
coefficients(lm_aircraft)['aircraft'] > 0
summary(lm_aircraft)$r.squared
table2 <- matrix(data = c('aircraft', 
                          'duration',
                          'passengers',
                          'ground speed',
                          'air speed',
                          'height',
                          'pitch',
                          lmp(lm_aircraft),
                          lmp(lm_duration),
                          lmp(lm_psg),
                          lmp(lm_gs),
                          lmp(lm_as),
                          lmp(lm_ht),
                          lmp(lm_pitch),
                          if_else(coefficients(lm_aircraft)['aircraft'] > 0, 'positive', 'negative'),
                          if_else(coefficients(lm_duration)['duration'] > 0, 'positive', 'negative'),
                          if_else(coefficients(lm_psg)['no_pasg'] > 0, 'positive', 'negative'),
                          if_else(coefficients(lm_gs)['speed_ground'] > 0, 'positive', 'negative'),
                          if_else(coefficients(lm_as)['speed_air'] > 0, 'positive', 'negative'),
                          if_else(coefficients(lm_ht)['height'] > 0, 'positive', 'negative'),
                          if_else(coefficients(lm_pitch)['pitch'] > 0, 'positive', 'negative')), 
                          nrow = 7, ncol = 3)
table2

```


###Step 14

Standardize variables and buidl table 3

```{r}
FAAmcs <- FAAmc
head(FAAmcs)
FAAmcs$aircraft <- (FAAmc$aircraft - mean(FAAmc$aircraft, na.rm = TRUE))/sd(FAAmc$aircraft, na.rm = TRUE)
FAAmcs$duration <- (FAAmc$duration - mean(FAAmc$duration, na.rm = TRUE))/sd(FAAmc$duration, na.rm = TRUE)
FAAmcs$no_pasg <- (FAAmc$no_pasg - mean(FAAmc$no_pasg, na.rm = TRUE))/sd(FAAmc$no_pasg, na.rm = TRUE)
FAAmcs$speed_ground <- (FAAmc$speed_ground - mean(FAAmc$speed_ground, na.rm = TRUE))/sd(FAAmc$speed_ground, na.rm = TRUE)
FAAmcs$speed_air <- (FAAmc$speed_air - mean(FAAmc$speed_air, na.rm = TRUE))/sd(FAAmc$speed_air, na.rm = TRUE)
FAAmcs$height <- (FAAmc$height - mean(FAAmc$height, na.rm = TRUE))/sd(FAAmc$height, na.rm = TRUE)
FAAmcs$pitch <- (FAAmc$pitch - mean(FAAmc$pitch, na.rm = TRUE))/sd(FAAmc$pitch, na.rm = TRUE)
head(FAAmcs)

lm_s_aircraft <- lm(distance ~ aircraft, data = FAAmcs)
lm_s_duration <- lm(distance ~ duration, data = FAAmcs)
lm_s_psg <- lm(distance ~ no_pasg, data = FAAmcs)
lm_s_gs <- lm(distance ~ speed_ground, data = FAAmcs)
lm_s_as <- lm(distance ~ speed_air, data = FAAmcs)
lm_s_ht <- lm(distance ~ height, data = FAAmcs)
lm_s_pitch <- lm(distance ~ pitch, data = FAAmcs)

table3 <- matrix(data = c('aircraft', 
                          'duration',
                          'passengers',
                          'ground speed',
                          'air speed',
                          'height',
                          'pitch',
                          coefficients(lm_s_aircraft)['aircraft'],
                          coefficients(lm_s_duration)['duration'],
                          coefficients(lm_s_psg)['no_pasg'],
                          coefficients(lm_s_gs)['speed_ground'],
                          coefficients(lm_s_as)['speed_air'],
                          coefficients(lm_s_ht)['height'],
                          coefficients(lm_s_pitch)['pitch'],
                          if_else(coefficients(lm_s_aircraft)['aircraft'] > 0, 'positive', 'negative'),
                          if_else(coefficients(lm_s_duration)['duration'] > 0, 'positive', 'negative'),
                          if_else(coefficients(lm_s_psg)['no_pasg'] > 0, 'positive', 'negative'),
                          if_else(coefficients(lm_s_gs)['speed_ground'] > 0, 'positive', 'negative'),
                          if_else(coefficients(lm_s_as)['speed_air'] > 0, 'positive', 'negative'),
                          if_else(coefficients(lm_s_ht)['height'] > 0, 'positive', 'negative'),
                          if_else(coefficients(lm_s_pitch)['pitch'] > 0, 'positive', 'negative')), 
                          nrow = 7, ncol = 3)
table3
```

###Step 15

Comparing the tables

Results between tables 1, 2, and 3 look to be consistent.
```{r}
table1_1
table2
table3

table0 <-  matrix(data = c('aircraft', 
                          'duration',
                          'passengers',
                          'ground speed',
                          'air speed',
                          'height',
                          'pitch', 
                          3,
                          6,
                          7,
                          2,
                          1,
                          4,
                          5), 
                  ncol = 2, nrow = 7)
table0
```

The final table shows the proper ranking of which variables look to be most important to the landing distance prediction.

##Check collinearity 

###Step 16

Comparing set regressions

```{r}
summary(lm_gs)
summary(lm_as)
lm_speed <- lm(distance~ speed_air + speed_ground, data = FAAmc)
summary(lm_speed)
cor.test(FAAmc$speed_ground, FAAmc$speed_air, method = 'pearson')
```

Based on the two individual models and the combined model, it looks that air speed is more highly suited as a variable. It is more significant, and the adjusted R-square is higher. When combining the two variables into a single model, ground speed becomes insignificant. These are highly correlated variables when using a pearson test, so there is certainly a degree of multicollinearity when including both.

##Variable selection based on our ranking in Table 0. 

###Step 17

Fitting additive models

```{r}
model1 <- lm_as
model2 <- lm(distance ~ speed_air + speed_ground, data = FAAmc)
model3 <- lm(distance ~ speed_air + speed_ground + aircraft, data = FAAmc)
model4 <- lm(distance ~ speed_air + speed_ground + aircraft + height, data = FAAmc)
model5 <- lm(distance ~ speed_air + speed_ground + aircraft + height + pitch, data = FAAmc)
model6 <- lm(distance ~ speed_air + speed_ground + aircraft + height + pitch + duration, data = FAAmc)
model7 <- lm(distance ~ speed_air + speed_ground + aircraft + height + pitch + duration + no_pasg, data = FAAmc)

rsq <- c(summary(model1)$r.squared,
         summary(model2)$r.squared,
         summary(model3)$r.squared,
         summary(model4)$r.squared,
         summary(model5)$r.squared,
         summary(model6)$r.squared,
         summary(model7)$r.squared
         )
plot(rsq, xlab = "number of variables")
```

As you can see, the graph shows how when the number of variables increases, the R squared value (and thus the relative determination of the prediction model) increases. However, after 4 variables, the increase is negligible.

###Step 18

Using Adjusted R Squared values

```{r}
arsq <- c(summary(model1)$adj.r.squared,
         summary(model2)$adj.r.squared,
         summary(model3)$adj.r.squared,
         summary(model4)$adj.r.squared,
         summary(model5)$adj.r.squared,
         summary(model6)$adj.r.squared,
         summary(model7)$adj.r.squared
         )
plot(arsq, xlab = "number of variables")
```

The trend is similar to the trend that was observed with the regular R squared values.

###Step 19

Using AIC values

```{r}
aic <- c(AIC(model1),
         AIC(model2),
         AIC(model3),
         AIC(model4),
         AIC(model5),
         AIC(model6),
         AIC(model7)
         )
plot(aic, xlab = "number of variables")
```

After the first two values, AIC dips a lot. Then after the third variable, AIC dips again. The dip between 4 and 5 is minimal, as is the dip betweeen 6 and 7 variables.

###Step 20
For a predictive model for LD, I would use the first four variables: air speed, ground speed, aircraft, and height. After that, the AIC dips too low, and there are not significant improvements to adjusted R squared beyond that. 

##Variable selection based on automate algorithm.

###Step 21
```{r warning = FALSE, error = FALSE}
library(MASS)
stepAIC(model5)
```

Using this StepAIC function in R, we can see that the AICs are insignificantly higher for 3 variables, 4, and 5. These are lower AIC values than the ones calculated in Step 19.
